Well-quasi-orderings on word languages \\ main edb3a6bc24be4a12fe4f383c75384a6dff9d1a3c \\ 2025-10-02 17:15:30 +0200

### 2025-10-06 src/infixes.tex
Build for lncs

\begin{theorem}
    \label{bounded-language:thm}
    Let $L$ be a \kl{bounded language} of $\Sigma^*$. Then,
    $L$ is a \kl{well-quasi-order} when endowed with the 
    \kl{infix relation} if and only if it included in a finite union of 
    products $S_i \cdot P_i$ where 
    $S_i$ is a \kl{chain} for the \kl{suffix relation}, and 
    $P_i$ is a \kl{chain} for the \kl{prefix relation},
    for all $1 \leq i \leq n$.
\end{theorem}

Let us first remark that if $S$ is a \kl{chain} for the \kl{suffix relation}
and $P$ is a \kl{chain} for the \kl{prefix relation}, then $SP$ is
\kl{well-quasi-ordered} for the \kl{infix relation}. This proves the (easy)
right-to-left implication of \cref{bounded-language:thm}.

\AP In order to prove the (difficult) left-to-right implication of
\cref{bounded-language:thm}, we will rely heavily on the
combinatorics of periodic words. Let us recall that a non-empty word $w \in
\Sigma^+$ is \intro(word){periodic} with period $x \in \Sigma^*$ if there
exists a $p \in \Nat$ such that $w \infleq x^p$. The \intro{periodic length} of
a word $u$ is the minimal length of a word $x$ such that $u$ is an \kl{infix}
of $x^p$ for some $p \in \Nat$ and $x \in \Sigma^+$.

\begin{lemma}
    \label{periodic-infixes:lem}
    Let $u,v \in \Sigma^*$ be two (non-empty) \kl{periodic words}
    having \kl{periodic lengths} $p$ and $q$ respectively.
    Then, if $u \infleq v$ and $\card{u} \geq \factorial[p]{p \times q}$,
    then $u$ and $v$ share the same \kl{periodic length}
    $p = q$.
\end{lemma}
\begin{proof}
    The fact that $u$ and $v$ are \kl{periodic length}
    respectively $p$ and $q$ translates into the fact that $u_{i+p} = u_i$ and
    $v_{i+q} = v_i$ for all indices $i \in \Nat$ such that those letters are
    well-defined.

    Now, assume that $u$ is an \kl{infix} of $v$, this provides the existence
    of a $k \in \Nat$ such that $u = v_{k} \cdots v_{k + \card{u} - 1}$. In
    particular, $v_{k+i+p} = v_{k+i}$ for all $i \in \Nat$ such that $k+i+p < k
    + \card{u}$. Since we also have $v_{k+i+q} = v_{k+i}$ for all $1 \leq i
    \leq \card{v} - k - q$. We conclude that both $u$ and $v$ are of
    \kl{periodic length} the greatest common divisor of $p$ and $q$, and by
    minimality of $q$ this must be equal to $q$ and to $p$.
\end{proof}

@@

\begin{corollary}
    \label{bounded-wqo-dwclosed:cor}
    Let $L$ be a \kl{bounded language} of $\Sigma^*$. Then,
    $L$ is a \kl{well-quasi-order} when endowed with the
    \kl{infix relation} if and only if $\dwset[\infleq]{L}$ is.
\end{corollary}
\begin{proof}
    Because $L \subseteq \dwset[\infleq]{L}$, the right-to-left implication
    is trivial.
    For the left-to-right implication, let us assume that $L$ is a
    \kl{well-quasi-ordered} language for the \kl{infix relation}.
    Then $L$ is included in a finite union 
    of products of \kl{chains}:
    \begin{equation*}
        L \subseteq \bigcup_{i = 1}^n S_i \cdot P_i \quad .
    \end{equation*}
    Remark that the \kl{downwards closure} of a product of two \kl{chains}
    is a finite union of products of two chains.
    As a consequence, we conclude that $\dwset[\infleq]{L}$ is itself included
    in a finite union of products of \kl{chains}.
    Note that this also proves that $\dwset[\infleq]{L}$ is a \kl{bounded language},
    hence that it is \kl{well-quasi-ordered} by the \kl{infix relation} 
    via
    \cref{bounded-language:thm}.
\end{proof}

One may think that all \kl{downwards closed} languages for the \kl{infix
relation} that are \kl{well-quasi-ordered} are \kl(language){bounded}. Note
that this is what happens in the case of the \kl{subword embedding}, where any
\kl{downwards closed} language for this relation is characterized by finitely
many excluded \kl{subwords}, hence provides a \kl{bounded language}. However, this
is not the case for the \kl{infix relation}, as we will now illustrate with the
following two examples.

\begin{example}
    \label{dwclosed-wqo-not-finite-excl:ex}
    Let $L \defined a^* b^* \cup b^* a^*$. This language is \kl{downwards
    closed} for the \kl{infix relation}, is \kl{well-quasi-ordered} for the
    \kl{infix relation}, but is characterized by an \emph{infinite} number 
    of excluded infixes, respectively of the form $ab^ka$ and $ba^kb$ where $k \geq 1$.
\end{example}

>>>>

\begin{theorem}[restate=bounded-language:thm,label=bounded-language:thm]
    \label{bounded-language:thm}
    Let $L$ be a \kl{bounded language} of $\Sigma^*$. Then,
    $L$ is a \kl{well-quasi-order} when endowed with the 
    \kl{infix relation} if and only if it included in a finite union of 
    products $S_i \cdot P_i$ where 
    $S_i$ is a \kl{chain} for the \kl{suffix relation}, and 
    $P_i$ is a \kl{chain} for the \kl{prefix relation},
    for all $1 \leq i \leq n$.
\end{theorem}

Let us first remark that if $S$ is a \kl{chain} for the \kl{suffix relation}
and $P$ is a \kl{chain} for the \kl{prefix relation}, then $SP$ is
\kl{well-quasi-ordered} for the \kl{infix relation}. This proves the (easy)
right-to-left implication of \cref{bounded-language:thm}. Furthermore, any
language $L$ included in a finite union of products $S_i \cdot P_i$
is in fact a \kl{bounded language}. \todo[inline]{I think we can in fact construct chains that are not bounded languages. E.g. $ab, aba^2b, aba^2ba^3b, \ldots$?}

\todo[inline]{aliaume: yes, this way nonsense.}

\AP In order to prove the (difficult) left-to-right implication of
\cref{bounded-language:thm}, we will rely heavily on the
combinatorics of periodic words. Let us recall that a non-empty word $w \in
\Sigma^+$ is \intro(word){periodic} with period $x \in \Sigma^*$ if there
exists a $p \in \Nat$ such that $w \infleq x^p$. The \intro{periodic length} of
a word $u$ is the minimal length of a period $x$ of $u$.

\begin{lemma}
    \label{periodic-infixes:lem}
    Let $u,v \in \Sigma^*$ be two (non-empty) \kl{periodic words}
    having \kl{periodic lengths} $p$ and $q$ respectively.
    Then, if $u \infleq v$ and $\card{u} \geq pq$,
    then $u$ and $v$ share the same \kl{periodic length}
    $p = q$.
\end{lemma}
\begin{proof}
    The fact that $u$ and $v$ are \kl{periodic length}
    respectively $p$ and $q$ translates into the fact that $u_{i+p} = u_i$ and
    $v_{i+q} = v_i$ for all indices $i \in \Nat$ such that those letters are
    well-defined.

    Now, assume that $u$ is an \kl{infix} of $v$, this provides the existence
    of a $k \in \Nat$ such that $u = v_{k} \cdots v_{k + \card{u} - 1}$. In
    particular, $v_{k+i+p} = v_{k+i}$ for all $i \in \Nat$ such that $k+i+p < k
    + \card{u}$. Due to the periodicity of $v$, we also have $v_{k+i+q} = v_{k+i}$ for all $i$ with $k + i + q \leq k + \card{u}$. We conclude that both $u$ and $v$ are of
    \kl{periodic length} the greatest common divisor of $p$ and $q$, and by
    minimality of $q$ this must be equal to $q$ and to $p$.
    \todo{maybe elaborate on why $pq$ as bound on $\card{u}$ (also technically $2\max\{p,q\}$ would be enough, but maybe more cumbersome?}
\end{proof}

\begin{figure}
	\begin{center}
		\input{fig/periodic-infixed.tex}
	\end{center}
	\caption{Alignment of two periodic words with periods of length $6$ and $4$, respectively. We can conclude that the periodic length is at most $2$.}
	\label{fig:periodic-infixes}
\end{figure}

@@

\begin{corollary}
    \label{bounded-wqo-dwclosed:cor}
    Let $L$ be a \kl{bounded language} of $\Sigma^*$. Then,
    $L$ is a \kl{well-quasi-order} when endowed with the
    \kl{infix relation} if and only if $\dwset[\infleq]{L}$ is.
\end{corollary}
\begin{proof}
    Because $L \subseteq \dwset[\infleq]{L}$, the right-to-left implication
    is trivial.
    For the left-to-right implication, let us assume that $L$ is a
    \kl{well-quasi-ordered} language for the \kl{infix relation}.
    Then $L$ is included in a finite union 
    of products of \kl{chains}: \todo{reference}
    \begin{equation*}
        L \subseteq \bigcup_{i = 1}^n S_i \cdot P_i \quad .
    \end{equation*}
    Remark that the \kl{downwards closure} of a product of two \kl{chains}
    is a finite union of products of two chains. \todo{reference/proof}.
    As a consequence, we conclude that $\dwset[\infleq]{L}$ is itself included
    in a finite union of products of \kl{chains}.
    Note that this also proves that $\dwset[\infleq]{L}$ is a \kl{bounded language},
    hence that it is \kl{well-quasi-ordered} by the \kl{infix relation} 
    via
    \cref{bounded-language:thm}.
\end{proof}

One may think that all \kl{downwards closed} languages for the \kl{infix
relation} that are \kl{well-quasi-ordered} are \kl(language){bounded}. Note
that this is what happens in the case of the \kl{subword embedding}, where any
\kl{downwards closed} language for this relation is characterized by finitely
many excluded \kl{subwords}, hence provides a \kl{bounded language} \todo{wrong? $\{a,b\}^*$ is downward closed but not bounded}. However, this
is not the case for the \kl{infix relation}, as we will now illustrate with the
following two examples.

\begin{example}
    \label{dwclosed-wqo-not-finite-excl:ex}
    Let $L \defined a^* b^* \cup b^* a^*$ \todo{However, this language is bounded, so this seems to be a non sequitur}. This language is \kl{downwards
    closed} for the \kl{infix relation}, is \kl{well-quasi-ordered} for the
    \kl{infix relation}, but is characterized by an \emph{infinite} number 
    of excluded infixes, respectively of the form $ab^ka$ and $ba^kb$ where $k \geq 1$.
\end{example}

### 2025-10-06 wqo-on-words.tex
new: cut intro!
Build for lncs

% We include the title and author information based on the 
% `paper-meta.yaml` file.

\title{Well-quasi-orderings on word languages}

\author{
Nathan Lhote\thanks{Aix-Marseille University}
 \and
Aliaume Lopez\thanks{University of Warsaw}
 \and
Lia Schütze\thanks{Max Planck Institute for Software Systems}
}

% For the date, we first check if the user has provided a date,
% and otherwise use the git meta inforamtion (if available).
\date{2025-07-01 13:26:43 +0200\footnote{99064007a226e55157ecfac20bbb40804687f3e2 -- branch main at git@github.com:AliaumeL/word-ordering-wqo.git}}

@@

\knowledge{notion}
 | kl-usage

% Now, we create the document itself.
\begin{document}
% Generate the title page
\maketitle
% Print the abstract
\begin{abstract}
    The set of finite words over a well-quasi-ordered set is itself well-quasi-ordered. This seminal result by Higman is a cornerstone of the theory of well-quasi-orderings and has found numerous applications in computer science. However, this result is based on a specific choice of ordering on words, the (scattered) subword ordering. In this paper, we describe to what extent other natural orderings (prefix, suffix, and infix) on words can be used to derive Higman-like theorems. More specifically, we are interested in characterizing \emph{languages} of words that are well-quasi-ordered under these orderings. We show that a simple characterization is possible for the prefix and suffix orderings, and that under extra regularity assumptions, this also extends to the infix ordering. We furthermore provide decision procedures for a large class of languages, that contains regular and context-free languages.
    \paragraph{Keywords:}
    equivariant ideal, Hilbert basis, ideal membership problem, orbit finite, oligomorphic, well-quasi-ordering
\paragraph{Repository:} \repositoryUrl
\end{abstract}

\klogo\ This document uses \href{https://ctan.org/pkg/knowledge}{knowledge}:
\kl[kl-usage]{notion} points to its \intro[kl-usage]{definition}.

>>>>

@@

\begin{document}
%
\title{Well-quasi-orderings on word languages \\
       main edb3a6bc24be4a12fe4f383c75384a6dff9d1a3c \\
       2025-10-02 17:15:30 +0200
}
\titlerunning{WQO on Words}

\author{
        Nathan Lhote\inst{2}\orcidID{0000-0003-3303-5368} \and
        Aliaume Lopez\inst{3}\orcidID{0000-0002-4205-327X} \and
        Lia Schütze\inst{1}\orcidID{0000-0003-4002-5491}}
\authorrunning{N. Lhote and A. Lopez and L. Schütze}
\institute{
Max Planck Institute for Software Systems
 \and
Aix-Marseille University
 \and
INP Bordeaux, LaBRI, CNRS
}

\date{\today}

%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
    The set of finite words over a well-quasi-ordered set is itself well-quasi-ordered. This seminal result by Higman is a cornerstone of the theory of well-quasi-orderings and has found numerous applications in computer science. However, this result is based on a specific choice of ordering on words, the (scattered) subword ordering. In this paper, we describe to what extent other natural orderings (prefix, suffix, and infix) on words can be used to derive Higman-like theorems. More specifically, we are interested in characterizing \emph{languages} of words that are well-quasi-ordered under these orderings, and explore their properties and connections with other language theoretic notions. We furthermore give decition procedures when the languages are given by various comupational models such as automata, context-free grammars, and automatic structures.
\end{abstract}
